---
title: | 
  | Data analysis report - 
  | The effect of bariatric surgery on gravitational loading and its impact on bone mass
author: "Lucas Veras"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: true
    highlight: tango
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r source, eval = TRUE, echo = FALSE, include = FALSE}
library(here)
source(here("code/analysis.R"))
```

<br> 

This report contains the data analysis description of the manuscript entitled "The effect of bariatric surgery on gravitational loading and its impact on bone mass". All code contained in theses analysis is in a [github repository](https://github.com/verasls/BaSEIB_bariatric_surgery_bone_health). The [code](https://github.com/verasls/BaSEIB_bariatric_surgery_bone_health/tree/main/code) directory contains the [analysis.R](https://github.com/verasls/BaSEIB_bariatric_surgery_bone_health/blob/main/code/analysis.R) script, with all data analysis for the manuscript results, and the [functions](https://github.com/verasls/BaSEIB_bariatric_surgery_bone_health/tree/main/code/functions) subdirectory with function definitions. Also, the [.Rmd](https://github.com/verasls/BaSEIB_bariatric_surgery_bone_health/blob/main/report.Rmd) file used to generate this report is in the project root directory and the code to generate all figures included in the manuscript is inside the [figures](https://github.com/verasls/BaSEIB_bariatric_surgery_bone_health/tree/main/figures) directory.

<br> 

# Load and tidy data

Data was stored in a .csv file inside the `data/` directory and then read into a `tibble` by the [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) function. Then the categorical variables (`time`, `surgery`, `sex`, `diabetes`, `thiazides`, `smoker` and `menopause`) were all converted to `factor` and the `gravitational_loading` variable had its unit converted from Newton (N) to kilonewton (kN).

```{r load_data}
df <- read_csv(here("data/df.csv")) %>%
  mutate(
    across(
      c(time, surgery, sex, diabetes, thiazides, smoker, menopause),
      as.factor
    ),
    # Transform from N to kN
    gravitational_loading = gravitational_loading / 1000
  )
```

<br>

# Baseline characteristics

Baseline descriptive statistics were computed to characterize the sample. First, data from `time == 1` (pre-bariatric surgery assessment) were separated and the mean and standard deviation of the numerical variables (`age` and `BMI`) were calculated. After that, the proportions of the categorical variables were also computed, using the [`map()`](https://purrr.tidyverse.org/reference/map.html) function to apply `prop.table()` to all vectors containing the factors.

```{r descriptive}
df_baseline <- df %>%
  filter(time == 1) %>%
  select(age, sex, BMI, diabetes, thiazides, smoker, menopause)

# Numeric vars
df_baseline %>%
  summarise(
    across(where(is.double), list(mean = mean, sd = sd))
  )

# Categorical vars
cat_vars <- df_baseline %>%
  select(where(is.factor)) %>%
  names()

map(cat_vars, ~ prop.table(table(df_baseline[[.x]]))) %>%
  set_names(cat_vars)
```

<br>

# Linear mixed models: Time effect

All outcomes included in the manuscript had their time effect evaluated by linear mixed models using the `lmer()` function of the [`lme4`](https://cran.r-project.org/web/packages/lme4/index.html) package. These outcomes included bone mineral density, body composition, bone metabolism regulators, gravitational loading, sedentary behavior and physical activity intensities. The gravitational loading related variables were computed through a custom software written in Python and R, which was briefly described [elsewhere](https://github.com/verasls/daily_GRF).

All models and the information extracted from them were recursively created with the aid of the functional programming tools from the [`purrr`](https://purrr.tidyverse.org) package. An example of this process is shown below for the bone mineral density variables.

First, a character vector with the variable names was created in order to recursively build the model formula for each variable. All models had the same structure: the `time` variable (4 levels; pre-bariatric surgery, 1-month, 6-months and 12-months after bariatric surgery) was used as a fixed factor and the `subj` variable (the subjects) as a random factor. Therefore, the only varying part was the outcome variable and the general model formula was `outcome ~ time + (1 | subj)`. Then, all models were created applying the `lmer()` function to a list of formulas, and the models were also stored in a list. Afterwards, the estimated marginal means for all the `time` levels were calculated with the `emmeans()` function of the [`emmeans`](https://cran.r-project.org/web/packages/emmeans/index.html) package and the pairwise comparisons with Holm adjustment were computed with the `pairs()` method for an `emmGrid` object. Finally, the estimated marginal means were put into data frames and exported as .csv files to be used to build the plots.

```{r time_effect}
BMD_vars <- c("pc_TH_BMD", "pc_FN_BMD", "pc_LS_BMD", "pc_TR_BMD")
BMD_formula <- map(BMD_vars, ~ as.formula(paste0(.x, " ~ time + (1 | subj)")))
BMD_models <- map(BMD_formula, lmer, data = df)  %>% set_names(BMD_vars)
BMD_fixed_effects <- map(BMD_models, anova, type = 3, test = "F")
BMD_time_emm <- map(BMD_models, ~ emmeans(.x, ~ time))
BMD_pairwise <- map(BMD_time_emm, pairs, adjust = "holm")
# Write estimated marginal means to a file
list(
  data = map(BMD_time_emm, as.data.frame),
  name = map(BMD_vars, ~ paste0("emm_", .x, "_time.csv"))
) %>%
  pwalk(write_output)
```

<br>

# Linear mixed models: Prediction of bone mineral density changes

The influence of potential predictors of bone mineral density changes occurred during the first year after bariatric surgery was also tested. Prediction models were performed through mixed-models analysis, again using the `lmer()` function of the `lme4` package, in which predictors were setup as fixed factor while subjects and time were setup as random factors.

In these models, all variables were centered previously to entering the model.

```{r center_vars}
vars <- c(
  "TH_BMD", "FN_BMD", "LS_BMD",
  "whole_body_total_mass", "whole_body_fat_mass",
  "whole_body_lean_mass", "gravitational_loading"
)
centered_vars <- map(vars, ~ scale(df[[.x]])) %>%
  set_names(paste0(vars, "_centered")) %>%
  map(as.vector) %>%
  as_tibble
df <- cbind(df, centered_vars) %>% as_tibble
```
Similarly to the other models, functional programming was also applied to build the models and extract their information recursively. The name of the predictor variables were entered in a character vector and the model formulas were built using the general form `BMD_variable ~ predictor + (1 | subj) + (1 | time)`. Then the models were built and stored into a list, and their parameter estimates were extracted from the `coefficients` part of the `summary()` function output. Finally, the marginal R^2^ of each model was calculated using the `r2()` function of the [`lvmisc`](https://lveras.com/lvmisc/) package. An example of the models created with total hip bone mineral density as outcome is shown below.

```{r prediction_models}
predictors <- c(
  "whole_body_total_mass_centered", "whole_body_fat_mass_centered",
  "whole_body_lean_mass_centered", "gravitational_loading_centered"
)

# Total hip
TH_formula <- map(
  predictors,
  ~ as.formula(paste0("TH_BMD_centered ~ ", .x, " + (1 | subj) + (1 | time)"))
)
TH_models <- map(TH_formula, lmer, data = df) %>%
  set_names(predictors)
TH_estimates <- map(TH_models, summary) %>% map(coefficients)
TH_r2 <- map(TH_models, r2)
```
<br>

# R session info

R session info with all packages loaded and their versions.

```{r session_info, eval = TRUE}
sessioninfo::session_info()
```
